Intelligence and Age at First Intercourse: Cause or Confound?

### 2016-09-24 Common/content/writing/current_study.tex
update

We made the following predictions, based primarily upon \citet{harden2011don}:\\ 
Between Families,\\
1.\hypothesis{btw_2_2} Does Generation 2 (\ie, children's) intelligence predict Generation 2 AFI?: We expect intelligence to be associated with age of first intercourse because there is a sizable body of literature reporting that result \citep{kirby2002effective, manlove1998influence, raffaelli2003sexual, rodgers1994df}.\\ 
2.\hypothesis{btw_1_2} Does Generation 1 (\ie, mother's) intelligence predict Generation 2 AFI?: We also expect maternal intelligence to be associated with age of first intercourse because the heritability of intelligence is moderate-to-high \citep{Bouchard2004,devlin1997heritability}. If intelligence does causally influence AFI we would expect that the cross-generational association between AFI and intelligence would be weaker, but existent. However, if the intelligence-AFI relationship is the product of between-family confounds, then we would expect that the cross-generational association between AFI and intelligence would be stronger than the within-generation association because maternal intelligence would be more closely linked with household SES and various parental causes. In other words, maternal intelligence can serve as a proxy for many of the between-family confounds that are of concern in the current study. Comparably sized effects would also be consistent with a between-family confound. Given that \citet{harden2011don} found no within-family effect for intelligence, we expect that maternal intelligence will have a comparable or larger effect on between-family AFI than child intelligence.\\
Within Families,\\
3.\hypothesis{wth_2_2} Does Generation 2 intelligence predict Generation 2 AFI?: We do not expect to find within-family differences in intelligence correlating with AFI, given that \citet{harden2011don} did not report an effect.\\
4.\hypothesis{wth_1_2} Does Generation 1 intelligence predict Generation 2 AFI?: Unknown: it is possible that maternal intelligence will have an effect, as such a link would explain the between-family effects as well as many of the alternative household-level influences.\\
5.\hypothesis{btw_v_wth} Is the relationship consistent across cross-sectional and within-family designs?: Doubtful, we do not expect the results to be consistent across methods because both \citet{harden2011don} and \citet{Meredith2013} found no within-family effect, whereas the traditional findings from between-family studies find an effect \citep{kirby2002effective, manlove1998influence, raffaelli2003sexual}. The birth order literature is one example in which apparent causal links mitigate or disappear when variance is properly attributed to within- versus between-family sources.

>>>>

We made the following predictions, based primarily upon \citet{harden2011don}:\\ 
Between Families,\\
1.\hypothesis{btw_2_2} Does Generation 2 (\ie, children's) intelligence predict Generation 2 AFI?: We expect intelligence to be associated with age of first intercourse because there is a sizable body of literature reporting that result \citep{kirby2002effective, manlove1998influence, raffaelli2003sexual, rodgers1994df}.\\ 
2.\hypothesis{btw_1_2} Does Generation 1 (\ie, mother's) intelligence predict Generation 2 AFI?: We also expect maternal intelligence to be associated with age of first intercourse because the heritability of intelligence is moderate-to-high \citep{Bouchard2004,devlin1997heritability}. If intelligence does causally influence AFI, we would expect that the cross-generational association between AFI and intelligence would be weaker, but existent. However, if the intelligence-AFI relationship is the product of between-family confounds, then we would expect that the cross-generational association between AFI and intelligence would be stronger than the within-generation association because maternal intelligence would be more closely linked with household SES and various parental causes. In other words, maternal intelligence can serve as a proxy for many of the between-family confounds that are of concern in the current study. Comparably sized effects would also be consistent with a between-family confound. Given that \citet{harden2011don} found no within-family effect for intelligence, we expect that maternal intelligence will have a comparable or larger effect on between-family AFI than child intelligence.\\
Within Families,\\
3.\hypothesis{wth_2_2} Does Generation 2 intelligence predict Generation 2 AFI?: We do not expect to find within-family differences in intelligence correlating with AFI, given that \citet{harden2011don} did not report an effect.\\
4.\hypothesis{wth_1_2} Does Generation 1 intelligence predict Generation 2 AFI?: Unknown -- it is possible that maternal intelligence will have an effect, as such a link would explain the between-family effects as well as many of the alternative household-level influences.\\
5.\hypothesis{btw_v_wth} Is the relationship consistent across cross-sectional and within-family designs?: Doubtful, we do not expect the results to be consistent across methods because both \citet{harden2011don} and \citet{Meredith2013} found no within-family effect, whereas the traditional findings from between-family studies find an effect \citep{kirby2002effective, manlove1998influence, raffaelli2003sexual}. The birth order literature is one example in which apparent causal links mitigate or disappear when variance is properly attributed to within- versus between-family sources.

### 2016-09-24 Common/content/writing/design.tex
update

In this model, the relative difference in kin outcomes ($\mathrm{Y_{\Delta}}$ \eg, AFI) is predicted from the mean level of Y ($\mathrm{\bar{Y}}$, \eg mean AFI), the mean level of X ($\mathrm{\bar{X}}$ \eg, intelligence), and the between-kin intelligence difference ($\mathrm{X_{i\Delta}}$). The mean levels, reflecting between-family variance, support causal inference through at least partial control for genes and shared environment in previous generations. Within this model, there is explicit separation of within-family variance (with $\mathrm{Y_{\Delta}}$ and $\mathrm{X_{\Delta}}$), and between-family variance (with $\mathrm{\bar{Y}}$ and $\mathrm{\bar{X}}$).

>>>>

In this model, the relative difference in kin outcomes ($\mathrm{Y_{\Delta}}$; \eg, AFI) is predicted from the mean level of Y ($\mathrm{\bar{Y}}$; \eg mean AFI), the mean level of X ($\mathrm{\bar{X}}$; \eg, intelligence), and the between-kin intelligence difference ($\mathrm{X_{i\Delta}}$). The mean levels, reflecting between-family variance, support causal inference through at least partial control for genes and shared environment in previous generations. Within this model, there is explicit separation of within-family variance (with $\mathrm{Y_{\Delta}}$ and $\mathrm{X_{\Delta}}$), and between-family variance (with $\mathrm{\bar{Y}}$ and $\mathrm{\bar{X}}$).

### 2016-09-24 Common/content/writing/discussion.tex
update

Given our findings and those of \citet{harden2011don}, we find no evidence for intelligence being a direct causal influence on AFI. Instead, we direct future researchers to look at the general family environment,``little e,'' and other between-family factors correlated with maternal intelligence as likely causes of AFI.  Although ``Smart Teens Don't Have Sex (or Kiss Much Either)'' \citep{halpern2000smart} at a descriptive level, the specific and proximal causal factors do not appear to be much related to smartness.

>>>>

Given our findings and those of \citet{harden2011don}, we find no evidence for intelligence being a direct causal influence on AFI. Instead, we direct future researchers to look at the general family environment,``little e,'' and other between-family factors correlated with maternal intelligence as likely causes of AFI.  Although ``smart teens don't have sex (or kiss much either)'' \citep{halpern2000smart} at a descriptive level, their reasons for delaying these activities do not appear to be caused by their smartness.

### 2016-09-24 Common/content/writing/introduction.tex
update

Teenage sexual activity has interested academics in demography and psychology for many years \citep{Brooks-Gunn1989,kinsey1948sexual,santelli2000adolescent}. Anecdotal evidence from the popular media, (\eg, MTV's reality television franchise, \textit{16 and Pregnant}\nocite{mtv}), 
and academic research converge to document a relatively consistent secular decline in age at first intercourse  \citep[AFI; see][]{bozon2003,finer2007trends,Kann2014}. Early AFI is associated with downstream consequences, including lower educational attainment \citep{Harden2012,Spriggs2008,Wellings2001}, failure to meet education and career goals \citep{halpern2000smart}, increased risk of teenage pregnancy \citep{Leitenberg2000,Wellings2001}, and increased rates of sexually transmitted infections \citep[STIs;][]{kaestle2005young}. Moreover, beyond the obvious benefit of avoiding those negative outcomes, delaying AFI is associated with greater relationship satisfaction, perception of increased attractiveness, and higher household income \citep{Harden2012}. Because many of the negative consequences above are severe and long-reaching, it is important to identify the causal mechanisms associated with early AFI. One potential factor exerting causal influence on AFI is intelligence.

@@

Virtually all of the AFI-intelligence literature has used between-family designs. In analysis of data from such designs, a number of genetic and environmental influences, such as education and maternal intelligence, are confounded \citep{DOnofrio2013,harden2014genetic,Lahey2010,Rodgers2000}. By ignoring such confounds, the source of variance is ambiguous, and researchers that attribute the source to specific between- or within-family sources risk misattributions of causality \citep{Rowe1997,Rutter2007}. There is the potential for this type of confound in virtually all past research on the link between intelligence and AFI \citep{harden2011don,harden2014genetic,plomin2004intelligence,rodgers1999nature,rodgers1994df}. Thus, we need to critically evaluate whether intelligence has a causal influence on AFI or is rather a theoretically attractive confound. Sibling models have been used in the past to at least partially control for confounds that create causal ambiguity \citep[\eg][]{east1996younger,east1993sisters,geronimus1992socioeconomic,rodgers1990adolescent,rodgers1992sibling};  several of these studies used the sibling structure in earlier versions of the NLSY.  As in those early studies, we resolve some of these methodological challenges by using design innovations that emerge from the excellent cross-generational and longitudinal structure of the National Longitudinal Survey of Youth (NLSY; we use both the original NLSY79 survey and the NLSY-Children survey, described below).
%
\section{Cause or Confound?}
Numerous theories address the motivations for adolescents' initiation of first intercourse (see \citealp{Rodgers1996} or \citealp{Buhi2007} for reviews), and even more specific precursors to first intercourse \citep{Buhi2007,DOnofrio2010,kirby2002antecedents,miller1997timing,santelli1992risk}. Many of these theories emphasize biology/genetics, as adolescent pubertal development (and associated hormone changes) drives the onset of sexual behavior \citep{miller1999dopamine,udry1979age,udry1994nature}. Other theoretical frameworks use social/environmental processes to explain developing sexual involvement in adolescence, such as Social Learning \citep{diblasio1990adolescent,hogben1998using}, where social norms affect the likelihood of early sexual behavior; or Social Control theory \citep{hirschi2002causes}, where societal and cultural influences reduce the likelihood that individuals will act on their natural tendency toward sexual involvement. Under these environmental theories the underlying biology is typically ignored (\eg, Social Control theory), whereas under many of the biological/genetic theories, the environmental components are often ignored.

However, numerous articles have also advocated integrative models \citep[See][]{harden2008rethinking,harden2014genetic,udry1995sociology}. The integrative Biopsychosocial Model acknowledges both genetic and environmental contributions to human behavior \citep{Engel1977,petersen1987nature,rodgers1999nature}. Indeed, biology, psychology, and society/culture jointly influence adolescents' decisions to engage in sexual intercourse \citep{Meschke2000,zimmer2008ten}.
%
\subsection{Intelligence as a Cause}
The short-term risks of early AFI are primarily negative, whereas the rewards for delay are primarily positive. These consequences extend into adulthood -- early AFI has been related to adult delinquency \citep{harden2008rethinking}, anti-social behavior, and substance abuse \citep{boislard2011individual}, whereas those with delayed AFI have higher household incomes in adulthood \citep{Harden2012}. It is intuitively appealing to believe that intelligent individuals are more likely to observe this potential risk-reward tradeoff, and through volition act upon such observations by delaying first intercourse. Accordingly, intelligent individuals perceive the consequences of early AFI to negatively influence their lives and careers \citep{halpern2000smart,harden2011don}.

@@

However, \citet{halpern2000smart} and many of the other studies we have referenced \citep[e.g.,][]{mathews2009predictors,miller1997timing,Paul2000} used between-family, typically cross-sectional, designs\footnote{\citet{halpern2000smart} used the first wave of Add Health for most of their AFI-intelligence analyses.}. Such designs cannot logically distinguish between processes that act to create differences between families and processes that create differences among family members \citep{Lahey2010,Rodgers2000}. Thus the previous studies do not provide conclusive evidence that intelligence is the causal influence behind the AFI-intelligence relationship. Logically, other alternatives are that AFI has a causal link to intelligence (which is unlikely, for the obvious theoretical reasons, including that a child's intelligence precedes AFI in time) or that other confounds cause these two outcomes to correlate, but not causally. There are dozens, perhaps hundreds, of such confounds that can logically contend to explain the link between child intelligence and AFI; we review some of those confounds in the next section.
%
\subsection{Intelligence as a Confound}
An equally valid set of explanations exist in which intelligence is not the causal factor behind the AFI-intelligence relationship, but rather one of dozens of correlated potentially explanatory processes. Instead, various confounds including family-level selection effects, or third variables at the individual or family level could be causing the relationship. Between-family influences such as SES and maternal intelligence could drive the relationship. Socioeconomic status is associated with the onset of first intercourse \citep{Lammers2000}, which potentially explains many of the negative consequences linked with teenage pregnancy \citep{geronimus1992socioeconomic}, and is correlated with intelligence \citep{murray1998income,Neisser1996,Strenze2007}. Parental intelligence and parental education are also linked with child intelligence \citep{Bouchard2004,devlin1997heritability,mercy1982familial}, and pose viable alternative explanations in which parents could be influencing, or even actively  dissuading, their children from engaging in early intercourse. For example, daughters whose mothers communicated frequently about the risk associated with sexual intercourse were less likely to have unprotected sex and engaged in sex less frequently \citep{hutchinson2003role}. Thus it could be that intelligent mothers, not intelligent children, are the ones recognizing the consequences of early intercourse and acting accordingly. In order to better understand the causal relation between intelligence and AFI, we need to be able to untangle between- and within-family processes, using both data and designs that have the ability to separate these sources of variance.

@@

\subsection{Prior Within-Family Analyses} Two past studies have explicitly separated between- and within-family influences on the AFI-intelligence relationship \citep{harden2011don,nedelec2012exploring}. Harden and Mendle used 536 same-sex identical and fraternal twin pairs from the National Longitudinal Study of Adolescent to Adult Health Study (Add Health) to ``test[ ] whether relations between intelligence, academic achievement and age at first sex were due to unmeasured genetic and environmental differences between families.'' Twins who differed in their intelligence or their academic achievement did not differ in their age at first intercourse. They concluded that ''the association between intelligence and age at first sex could be attributed entirely to \textit{unmeasured environmental differences between families}.''(italics added, our own emphasis). Nedelec, Schwartz, Connolly, and Beaver (\citeyear{nedelec2012exploring}) conducted an exploratory analysis of MZ twin pairs from the same sample employed by Harden and Mendle, and \citet{halpern2000smart}, using intelligence difference scores to predict various social outcomes. They found consistent null results, though their samples were small and their statistical analyses were substantially underpowered.

>>>>

Teenage sexual activity has interested academics in demography, public health, and psychology for many years \citep{Brooks-Gunn1989,kinsey1948sexual,santelli2000adolescent}. Anecdotal evidence from the popular media, (\eg, MTV's reality television franchise, \textit{16 and Pregnant}\nocite{mtv}), and academic research converge to document a relatively consistent secular decline in age at first intercourse  \citep[AFI; see][]{bozon2003,finer2007trends,Kann2014}. Early AFI is associated with downstream consequences, including lower educational attainment \citep{Harden2012,Spriggs2008,Wellings2001}, failure to meet education and career goals \citep{halpern2000smart}, increased risk of teenage pregnancy \citep{Leitenberg2000,Wellings2001}, and increased rates of sexually transmitted infections \citep[STIs;][]{kaestle2005young}. Moreover, beyond the obvious benefit of avoiding those negative outcomes, delaying AFI is associated with greater relationship satisfaction, perception of increased attractiveness, and higher household income \citep{Harden2012}. Because many of the negative consequences above are severe and long-reaching, it is important to identify the causal mechanisms associated with early AFI. One potential factor exerting causal influence on AFI is intelligence.

@@

Virtually all of the AFI-intelligence literature has used between-family designs. In analysis of data from such designs, a number of genetic and environmental influences, such as education and maternal intelligence, are confounded \citep{DOnofrio2013,harden2014genetic,Lahey2010,Rodgers2000}. By ignoring such confounds, the source of variance is ambiguous, and researchers that attribute the source to specific between- or within-family sources risk misattributions of causality \citep{Rowe1997,Rutter2007}. There is the potential for this type of confound in virtually all past research on the link between intelligence and AFI \citep{harden2011don,harden2014genetic,plomin2004intelligence,rodgers1999nature,rodgers1994df}. Thus, we need to critically evaluate whether intelligence has a causal influence on AFI or is rather a theoretically attractive confound. Sibling models have been used in the past to at least partially control for confounds that create causal ambiguity \citep[\eg][]{east1996younger,east1993sisters,geronimus1992socioeconomic,rodgers1990adolescent,rodgers1992sibling};  several of these studies used the sibling structure in earlier versions of the NLSY.  As in those early studies, we resolve some of these methodological challenges by using design innovations that emerge from the excellent cross-generational and longitudinal structure of the National Longitudinal Survey of Youth (NLSY; we use both the original NLSY79 survey and the NLSY-Children survey, described later).
%
\section{Cause or Confound?}
Numerous theories address the motivations for adolescents' initiation of first intercourse (see \citealp{Rodgers1996} or \citealp{Buhi2007} for reviews), and even more specific precursors to first intercourse \citep{Buhi2007,DOnofrio2010,kirby2002antecedents,miller1997timing,santelli1992risk}. Many of these theories emphasize biology/genetics, as adolescent pubertal development (and associated hormone changes) drives the onset of sexual behavior \citep{miller1999dopamine,udry1979age,udry1994nature}. Other theoretical frameworks use social/environmental processes to explain developing sexual involvement in adolescence, such as Social Learning \citep{diblasio1990adolescent,hogben1998using}, where social norms affect the likelihood of early sexual behavior; or Social Control theory \citep{hirschi2002causes}, where societal and cultural influences reduce the likelihood that individuals will act on their natural tendency toward sexual involvement. Under these environmental theories the underlying biology is typically ignored, whereas under many of the biological/genetic theories, the environmental components are often ignored.

However, numerous articles have advocated integrative models \citep[See][]{harden2008rethinking,harden2014genetic,udry1995sociology}. The integrative Biopsychosocial Model acknowledges both genetic and environmental contributions to human behavior \citep{Engel1977,petersen1987nature,rodgers1999nature}. Indeed, biology, psychology, and society/culture jointly influence adolescents' decisions to engage in sexual intercourse \citep{Meschke2000,zimmer2008ten}.
%
\subsection{Intelligence as a Cause}
The short-term risks of early AFI are primarily negative, whereas the rewards for delay are primarily positive. These consequences extend into adulthood -- early AFI has been related to adult delinquency \citep{harden2008rethinking}, anti-social behavior, and substance abuse \citep{boislard2011individual}, whereas those with delayed AFI have higher household incomes in adulthood \citep{Harden2012}. It is intuitively appealing to believe that intelligent individuals are more likely to observe this potential risk-reward tradeoff, and through volition, act upon such observations by delaying first intercourse. Accordingly, intelligent individuals perceive the consequences of early AFI to negatively influence their lives and careers \citep{halpern2000smart,harden2011don}.

@@

However, \citet{halpern2000smart} and many of the other studies we have referenced \citep[e.g.,][]{mathews2009predictors,miller1997timing,Paul2000} used between-family -- typically cross-sectional -- designs.\footnote{\citet{halpern2000smart} used the first wave of Add Health for most of their AFI-intelligence analyses.} Such designs cannot logically distinguish between processes that act to create differences between families and processes that create differences among family members \citep{Lahey2010,Rodgers2000}. Thus the previous studies do not provide conclusive evidence that intelligence is the causal influence behind the AFI-intelligence relationship. Logically, other alternatives are that AFI has a causal link to intelligence (which is unlikely, for the obvious theoretical reasons, including that a child's intelligence precedes AFI in time) or that other confounds cause these two outcomes to correlate, but not causally. There are dozens, perhaps hundreds, of such confounds that can logically contend to explain the link between child intelligence and AFI; we review some of those confounds in the next section.
%
\subsection{Intelligence as a Confound}
An equally valid set of explanations exist in which intelligence is not the causal factor behind the AFI-intelligence relationship, but rather one of dozens of correlated potentially explanatory processes. Instead, various confounds including family-level selection effects, or third variables at the individual or family level could be causing the relationship. Between-family influences such as SES and maternal intelligence could drive the relationship. Socioeconomic status is associated with the onset of first intercourse \citep{Lammers2000}, which potentially explains many of the negative consequences linked with teenage pregnancy \citep{geronimus1992socioeconomic}, and is correlated with intelligence \citep{murray1998income,Neisser1996,Strenze2007}. Parental intelligence and parental education are also linked with child intelligence \citep{Bouchard2004,devlin1997heritability,mercy1982familial}, and pose viable alternative explanations in which parents could be influencing, or even actively  dissuading, their children from engaging in early intercourse. For example, daughters whose mothers communicated frequently about the risk associated with sexual intercourse were less likely to have unprotected sex and engaged in sex less frequently \citep{hutchinson2003role}. Thus it could be that intelligent mothers, not intelligent children, are the ones recognizing the consequences of early intercourse and acting accordingly. In order to better understand the causal relation between intelligence and AFI, we need to be able to untangle between- and within-family processes, using both data and designs that have the ability to separate these sources of variance.

@@

\subsection{Prior Within-Family Analyses} Two past studies have explicitly separated between- and within-family influences on the AFI-intelligence relationship \citep{harden2011don,nedelec2012exploring}. Harden and Mendle used 536 same-sex identical and fraternal twin pairs from the National Longitudinal Study of Adolescent to Adult Health Study (Add Health) to ``test[ ] whether relations between intelligence, academic achievement and age at first sex were due to unmeasured genetic and environmental differences between families.'' Twins who differed in their intelligence or their academic achievement did not differ in their age at first intercourse. They concluded that ''the association between intelligence and age at first sex could be attributed entirely to \textit{unmeasured environmental differences between families}''(italics added, our own emphasis). Nedelec, Schwartz, Connolly, and Beaver (\citeyear{nedelec2012exploring}) conducted an exploratory analysis of MZ twin pairs from the same sample employed by Harden and Mendle, and \citet{halpern2000smart}, using intelligence difference scores to predict various social outcomes. They found consistent null results, though their samples were small and their statistical analyses were substantially underpowered.

### 2016-09-24 Common/content/writing/results_between.tex
update

First, we examined the between-family results. We tested whether the family average of Gen2 AFI could be predicted by the family averages of Gen1 intelligence and of Gen2 intelligence. We evaluated the influences both independently and simultaneously. All intelligence scores have been standardized by generation ($\overline{g} = 0$, sd $= 1$), prior to averaging by household. AFI scores have been standardized by gender ($\overline{\mathrm{AFI}} = 0$, sd $= 1$), prior to averaging by household. In Tables \ref{table_Mean_Mom_Intelligence_Mean_Child_AFI_9} - \ref{table_Mean_Joint_Intelligence_Mean_Child_AFI_9}, we have reported results for three different linking methods:
\begin{itemize} 
\item The mixed model, which contains the firstborn child (both males and females) of each sister;
\item The daughters model, which contains the firstborn girls; and 
\item The sons model, which contains the firstborn sons).\end{itemize}
In the spirit of transparency, we've reported all three methods. However, because all three linking methods reported similar findings, we will focus the results section on the mixed model and only discuss the other two methods when they deviate.
\subsubsection{Gen1 Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Gen1 sister averages (NLSY79 mothers) of standardized AFQT scores were used to predict Gen2 averages of gender-standardized AFI. Table \ref{table_Mean_Mom_Intelligence_Mean_Child_AFI_9} displays the results by Gen2 categories. The mixed model reports the averages of the firstborn child (both males and females) of each maternal sister (n $= 342$). A one-unit increase in the average standardized intelligence of the children's mothers predicted a statistically significant increase of $.013$ standard deviations in average Gen2 AFI. When we transform the coefficient into a standardized beta weight ($\beta_{Gen1 Intell} = .299$), a standard-deviation increase in the averaged mother's intelligence predicts a .299 standard-deviation increase in the average of the cousins' AFI. The adjusted R$^{2}$ was .087.

\subsubsection{Gen2 Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Gen2 averages of standardized intelligence scores were used to predict Gen2 averages of gender-standardized AFI. Table \ref{table_Mean_Child_Intelligence_Mean_Child_AFI_9} displays the results by Gen2 categories. The mixed model reports the averages of the firstborns of each of the NLSY79 mothers (sisters) (n $= 344$). A one-unit increase in the average standardized intelligence of the children predicted a statistically significant $\approx .075$ standard-deviation increase in average Gen2 cousins' AFI. When we transform the coefficient into a standardized beta weight ($\beta_{Gen2 Intell} = .128$), a standard-deviation increase in the averaged cousins' intelligence predicts a .128 standard-deviation increase in the average of the cousins' AFI. The adjusted R$^{2}$ was $.014$.

\subsubsection{Joint Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Results from the Gen1 maternal sister averages of standardized AFQT scores and Gen2 averages of standardized intelligence scores predicting Gen2 averages of gender-standardized AFI are displayed in Table \ref{table_Mean_Joint_Intelligence_Mean_Child_AFI_9}. In the mixed model, Gen1 (maternal) intelligence was significantly associated with Gen2 AFI (p $< .01$), while Gen2 (child) intelligence was not significantly associated with Gen2 AFI. A one-unit increase in the average standardized intelligence of the children's mothers predicted $.013$ standard-deviation increase in average Gen2 AFI, after controlling for Gen2 cousin averages of standardized intelligence scores. When we transform the coefficients into a standardized beta weight ($\beta_{Gen1 Intell} = .303$; $\beta_{Gen2 Intell} = -.003$), a standard-deviation increase in the averaged mother's intelligence predicts a .303 standard-deviation increase in the average of the cousins' AFI. The adjusted R$^{2}$ was $.086$. The total variance explained by the Joint model ($R^{2}$ = 9.1$\%$) is nearly identical to the Gen1 model($R^{2}$=9$\%$). Gen2 intelligence explains an additional .1\% of the variance.

When we broaden our sample to all Mother-Child pairs, we see that the relationship between Gen 2 intelligence and Gen2 AFI is small $(r =.139)$, and smaller than the relationship between Gen1 intelligence and Gen2 AFI $(r=.215$; see Figure \ref{plot_gen2_afi}). \pagebreak
\begin{landscape}
\begin{longtable}{@{\extracolsep{5pt}}lccc} 
\caption{Between-Family: Gen1 Intelligence Predicts Gen2 AFI}\label{table_Mean_Mom_Intelligence_Mean_Child_AFI_9}
\\[-1.8ex]\hline 
\hline \\[-3.8ex] 
& \multicolumn{3}{c}{\textit{Dependent variable:} Average of Gen2 AFI} \\ 
\cline{2-4}
\partialinput{10}{21}{../Common/content/tables/table_Mean_Mom_Intelligence_Mean_Child_AFI_9.tex}\\[-7ex]
\textit{Notes:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\[2ex]
& \multicolumn{3}{r}{\parbox{.6\linewidth}{\footnotesize Gen1-sister averages (NLSY79 mothers) of standardized AFQT scores predict Gen2 averages of gender-standardized AFI.}} \\ 
\end{longtable}\pagebreak

@@

\begin{longtable}{@{\extracolsep{5pt}}lccc} 
\caption{Between-Family: Gen1 \& Gen2 Intelligence Predict Gen2 AFI}\label{table_Mean_Joint_Intelligence_Mean_Child_AFI_9}
\\[-1.8ex]\hline 
\hline \\[-3.8ex] 
& \multicolumn{3}{c}{\textit{Dependent variable:} Average of Gen2 AFI} \\ 
\cline{2-4}
\partialinput{10}{22}{../Common/content/tables/table_Mean_Joint_Intelligence_Mean_Child_AFI_9.tex}\\[-7ex]
\textit{Notes:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\[2ex]
& \multicolumn{3}{r}{\parbox{.6\linewidth}{\footnotesize Gen1-sister averages (NLSY79 mothers) of standardized AFQT scores and Gen2-cousin averages of standardized intelligence scores predict Gen2 averages of gender-standardized AFI.}} \\ 
\end{longtable}\pagebreak

>>>>

First, we examined the between-family results. We tested whether the family average of Gen2 AFI could be predicted by the family averages of Gen1 intelligence and of Gen2 intelligence. We evaluated the influences both independently and simultaneously. All intelligence scores have been standardized by generation ($\overline{g} = 0$, sd $= 1$), prior to averaging by household. AFI scores have been standardized by gender ($\overline{\mathrm{AFI}} = 0$, sd $= 1$), prior to averaging by household. In Tables \ref{table_Mean_Mom_Intelligence_Mean_Child_AFI_9} - \ref{table_Mean_Joint_Intelligence_Mean_Child_AFI_9}, we have reported results for three different linking methods:
\begin{itemize} 
\item The mixed model, which contains the firstborn child of each sister;
\item The daughters model, which contains the firstborn daughters; and 
\item The sons model, which contains the firstborn sons).\end{itemize}
In the spirit of transparency, we have reported all three methods. However, because all three linking methods reported similar findings, we will focus the results section on the mixed model and only discuss the other two methods when they deviate. We have also provided zero-order and pairwise semi-partial correlations for all between-family variables (Tables \ref{table_cor} and \ref{table_spcor_btw}), using the mixed-model data and the ppcor \R library \citep{kim2015ppcor}.

\subsubsection{Gen1 Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Gen1 sister averages (NLSY79 mothers) of standardized AFQT scores were used to predict Gen2 averages of gender-standardized AFI. Table \ref{table_Mean_Mom_Intelligence_Mean_Child_AFI_9} displays the results by Gen2 category. The mixed model reports the averages of the firstborn child (both males and females) of each maternal sister (n $= 342$). A one-unit increase in the average standardized intelligence of the children's mothers predicted a statistically significant increase of $.013$ standard deviations in average Gen2 AFI. When we transform the coefficient into a standardized beta weight ($\beta_{Gen1 Intell} = .299$), a standard-deviation increase in the averaged mother's intelligence predicts a .299 standard-deviation increase in the average of the cousins' AFI. The adjusted R$^{2}$ was .087.

\subsubsection{Gen2 Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Gen2 averages of standardized intelligence scores were used to predict Gen2 averages of gender\hyph standardized AFI. Table \ref{table_Mean_Child_Intelligence_Mean_Child_AFI_9} displays the results by Gen2 category. The mixed model reports the averages of the firstborns of each of the NLSY79 mothers (sisters) (n $= 344$). A one-unit increase in the average standardized intelligence of the children predicted a statistically significant $\approx .075$ standard-deviation increase in average Gen2 cousins' AFI. When we transform the coefficient into a standardized beta weight ($\beta_{Gen2 Intell} = .128$), a standard-deviation increase in the averaged cousins' intelligence predicts a .128 standard-deviation increase in the average of the cousins' AFI. The adjusted R$^{2}$ was $.014$.

\subsubsection{Joint Mean Intelligence $\rightarrow$ Gen2 Mean AFI} Results from the Gen1 maternal sister averages of standardized AFQT scores and Gen2 averages of standardized intelligence scores predicting Gen2 averages of gender\hyph standardized AFI are displayed in Table \ref{table_Mean_Joint_Intelligence_Mean_Child_AFI_9}. In the mixed model, Gen1 (maternal) intelligence was significantly associated with Gen2 AFI (p $< .01$), while Gen2 (child) intelligence was not significantly associated with Gen2 AFI. A one-unit increase in the average standardized intelligence of the children's mothers predicted $.013$ standard-deviation increase in average Gen2 AFI, after controlling for Gen2 cousin averages of standardized intelligence scores. When we transform the coefficients into standardized beta weights ($\beta_{Gen1 Intell} = .303$; $\beta_{Gen2 Intell} = -.003$), a standard-deviation increase in the averaged mother's intelligence predicts a .303 standard-deviation increase in the average of the cousins' AFI. The adjusted R$^{2}$ was $.086$. The total variance explained by the Joint model ($R^{2}$ = 9.1$\%$) is nearly identical to the Gen1 model($R^{2}$=9$\%$). Gen2 intelligence explains an additional .1\% of the variance.

When we broaden our sample to all Mother-Child pairs, we see that the relationship between Gen 2 intelligence and Gen2 AFI is small $(r =.139)$, and smaller than the relationship between Gen1 intelligence and Gen2 AFI $(r=.215$; see Figure \ref{plot_gen2_afi}).

\begin{landscape}
\begin{longtable}{@{\extracolsep{5pt}}lccc} 
\caption{Between-Family: Gen1 Intelligence Predicts Gen2 AFI}\label{table_Mean_Mom_Intelligence_Mean_Child_AFI_9}
\\[-1.8ex]\hline 
\hline \\[-3.8ex] 
& \multicolumn{3}{c}{\textit{Dependent variable:} Average of Gen2 AFI} \\ 
\cline{2-4}
\partialinput{10}{21}{../Common/content/tables/table_Mean_Mom_Intelligence_Mean_Child_AFI_9.tex}\\[-7ex]
\textit{Notes:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\[2ex]
& \multicolumn{3}{r}{\parbox{.6\linewidth}{\footnotesize Gen1-sister averages (NLSY79 mothers) of standardized AFQT scores predict Gen2 averages of gender-standardized AFI.}} \\ 
\end{longtable}\pagebreak

@@

\begin{longtable}{@{\extracolsep{5pt}}lccc} 
\caption{Between-Family: Gen1 \& Gen2 Intelligence Predict Gen2 AFI}\label{table_Mean_Joint_Intelligence_Mean_Child_AFI_9}
\\[-1.8ex]\hline 
\hline \\[-3.8ex] 
& \multicolumn{3}{c}{\textit{Dependent variable:} Average of Gen2 AFI} \\ 
\cline{2-4}
\partialinput{10}{22}{../Common/content/tables/table_Mean_Joint_Intelligence_Mean_Child_AFI_9.tex}\\[-7ex]
\textit{Notes:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\[2ex]
& \multicolumn{3}{r}{\parbox{.6\linewidth}{\footnotesize Gen1-sister averages (NLSY79 mothers) of standardized AFQT scores and Gen2-cousin averages of standardized intelligence scores predict Gen2 averages of gender-standardized AFI.}} \\ 
\end{longtable}
